---
title: "Getting Started with bml"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with bml}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

The `bml` package implements *Bayesian Multiple-Membership Multilevel Models with Parameterizable Weight Functions*. These models are designed for situations where higher-level outcomes depend on the combined influence of multiple lower-level units. Instead of aggregating lower-level data to higher levels (risking aggregation bias) or disaggregating outcomes to lower levels (artificially inflating sample size), `bml` explicitly models how lower-level units jointly shape higher-level outcomes.

## Installation

Install the stable version from CRAN:

```{r}
install.packages("bml")
```

You'll also need JAGS installed on your system. See the [installation vignette](installation.html) for details.

## Basic Example

Let's start with a simple example using coalition government data. Each coalition (higher-level unit) is composed of multiple political parties (lower-level units), and we want to model how party characteristics influence coalition outcomes.

```{r setup}
library(bml)
data(coalgov)

# Examine the data structure
head(coalgov[, c("gid", "pid", "pname", "n", "finance", "dur_wkb", "event_wkb")])
```

### Understanding the Data Structure

The `coalgov` dataset is in **long format** where each row represents a party's participation in a coalition:

- `gid`: Government (coalition) identifier
- `pid`: Party identifier
- `n`: Number of parties in the coalition
- `finance`: Party's financial dependence on member contributions (standardized)
- `dur_wkb`: Coalition duration in days
- `event_wkb`: Early termination indicator (1 = terminated early, 0 = censored)

### Model 1: Basic Multiple-Membership Model

Our first model examines coalition duration as a function of party-level characteristics, aggregated using equal weights:

```{r}
mod1 <- bml(
  Surv(dur_wkb, event_wkb) ~ 1 + majority +
    mm(
      id = id(pid, gid),
      vars = vars(finance),
      fn = fn(w ~ 1/n, c = TRUE),
      RE = TRUE
    ),
  family = "Weibull",
  data = coalgov,
  n.iter = 2000,
  n.burnin = 500,
  seed = 1
)

summary(mod1)
```

**Breaking down the formula:**

- `Surv(dur_wkb, event_wkb)`: Survival outcome (duration and event indicator)
- `majority`: Government-level covariate (binary indicator)
- `mm()`: Multiple-membership block containing:
  - `id = id(pid, gid)`: Party ID (member) and government ID (group)
  - `vars = vars(finance)`: Party-level covariate to aggregate
  - `fn = fn(w ~ 1/n, c = TRUE)`: Weight function (equal weights, constrained to sum to 1)
  - `RE = TRUE`: Include party-specific random effects

### Model 2: Parameterizing the Weight Function

A key feature of `bml` is the ability to model aggregation weights as a function of covariates. Instead of assuming all parties have equal influence, we can test whether seat share affects a party's weight in the coalition:

```{r}
mod2 <- bml(
  Surv(dur_wkb, event_wkb) ~ 1 +
    majority +
    mm(
      id = id(pid, gid),
      vars = vars(finance),
      fn = fn(w ~ 1 / (1 + (n - 1) * exp(-(b0 + b1 * pseat))), c = TRUE),
      RE = TRUE
    ),
  family = "Weibull",
  priors = c("b.w ~ dnorm(0, 1)"), # prior on the weight parameters
  data = coalgov,
  seed = 1,
  monitor = TRUE
)

summary(mod2)
```

**New elements:**

- The weight function now includes `pseat` (party seat share) with parameter `b1`
- `priors`: Specify prior for the weight function parameter
- `monitor = TRUE`: Save MCMC chains for diagnostics

**Interpretation:**

- If `b1 ≈ 0`: Seat share doesn't affect weights (equal influence)
- If `b1 > 0`: Larger parties have more weight
- If `b1 < 0`: Smaller parties have more weight

### Visualizing Results

```{r}
# Diagnostic plot for weight parameter
monetPlot(mod2, parameter = "b.w[1]", label = "Seat share effect")

# MCMC diagnostics
mcmcDiag(mod2, parameters = "b.w[1]")
```

## Next Steps

- See the [Model documentation](model.html) for mathematical details
- Explore [Examples](examples.html) for spatial, network, and aggregation applications
- Check the [FAQ](faq.html) for common questions
- Read the accompanying paper: [Rosche (2026)](https://osf.io/preprints/socarxiv/4bafr)

## Key Concepts

- **Multiple-membership structure:** Higher-level units (coalitions) contain multiple lower-level units (parties), and lower-level units can appear in multiple higher-level units.
- **Weight functions:** Control how lower-level characteristics are aggregated. Can be:
  - Fixed (e.g., equal weights `w ~ 1/n`)
  - Endogenous (estimated from data, e.g., `w ~ 1/n^exp(b1*x)`)
- **Constraint parameter `c`:**
  - `c = TRUE`: Weights sum to 1 within each group
  - `c = FALSE`: No constraint on weight sums
- **Random effects:** Account for unobserved heterogeneity at the member level (`RE = TRUE` in `mm()`) or nesting level (`type = "RE"` in `hm()`)
