---
title: "Installation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Installation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`bml` fits Bayesian multiple-membership multilevel models via **JAGS**, so installation has two parts:

1.  Install JAGS (system software)
2.  Install the `bml` R package (from CRAN or GitHub)

## 1. Install JAGS

Install the latest JAGS version for your operating system from:

-   <http://mcmc-jags.sourceforge.net/>

After installing JAGS, restart R so `rjags` can find it.

## 2. Install `bml` R package

### Option A: Install from CRAN (Recommended)

The stable release version is available on CRAN:

```{r, eval = FALSE}
install.packages("bml")
```

### Option B: Install development version from GitHub

For the latest development features, install from GitHub using `{remotes}`:

```{r, eval = FALSE}
install.packages("remotes")
remotes::install_github("benrosche/bml")
```

If you want to build vignettes locally during installation:

```{r, eval = FALSE}
remotes::install_github("benrosche/bml", build_vignettes = TRUE)
```

## 3. Verify the installation

```{r, eval = FALSE}
library(bml)
packageVersion("bml")

# Optional: confirm that rjags is linked against JAGS
rjags::jags.version()
```

## Troubleshooting

**`Error: (converted from warning) ...` while installing**

If `{remotes}` treats an installation warning as an error on your machine, you can opt out:

```{r, eval = FALSE}
Sys.setenv(R_REMOTES_NO_ERRORS_FROM_WARNINGS = "true")
remotes::install_github("benrosche/bml")
```

**JAGS not found / `rjags` fails to load**

-   Make sure JAGS is installed (not just the R packages), then restart R.
-   On Windows, confirm you installed the 64-bit JAGS build if you are using 64-bit R.
